\input{../LecturesCommon/preamble.tex}
% set new colors according to AMU design
% http://dircom.univ-amu.fr/sites/dircom.univ-amu.fr/files/charte_graphique_generale_amu_0.pdf
\definecolor{AMUDarkBlue}{RGB}{0,103,186}
\definecolor{AMULightBlue}{RGB}{110,188,253}
\definecolor{AMUYellow}{RGB}{246,189,23}
%\usecolortheme[named=AMUDarkBlue]{structure}
% top bar with section navigation
%\setbeamercolor{palette tertiary}{bg=black,fg=white}
% 2nd top bar with subsection navigation (only if subsection=true)
%\setbeamercolor{palette secondary}{bg=AMUDarkBlue!50,fg=white}
% title bar of each slide
%\setbeamercolor{titlelike}{bg=AMUDarkBlue,fg=white}
%\setbeamercolor{title}{bg=AMUDarkBlue,fg=white}
%\setbeamercolor{subtitle}{fg=white}
\setbeamercolor{block title}{bg=AMUDarkBlue,fg=white}
\setbeamercolor{block body}{bg=AMUDarkBlue!30}
\setbeamercolor{block title example}{bg=AMUYellow,fg=white}
\setbeamercolor{block body example}{bg=AMUYellow!30}
\setbeamercolor{block title alerted}{bg=red,fg=white}
\setbeamercolor{block body alerted}{bg=red!30}
% \setbeamercolor{colorbox}{bg=HUSand!30}
% http://tex.stackexchange.com/questions/74303
\usepackage{etoolbox}
\AtBeginEnvironment{exampleblock}{%
  \setbeamercolor{itemize item}{fg=AMUYellow!70}
}
\AtBeginEnvironment{alertblock}{%
  \setbeamercolor{itemize item}{fg=red!70}
}
\AtBeginEnvironment{block}{%
  \setbeamercolor{itemize item}{fg=AMUDarkBlue!70}
}
% \rowcolors{2}{Gray!30}{white}

% apply default template partials
\useinnertheme{rectangles}
\useoutertheme[subsection=false]{miniframes}
\setbeamertemplate{navigation symbols}{}
% reset the footer: show slide numbers
\setbeamertemplate{footline}{%
  \begin{beamercolorbox}[colsep=1.5pt]{upper separation line foot}
  \end{beamercolorbox}
  \begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,%
    leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
    {\usebeamerfont{title in head/foot}\insertshorttitle}%
    \,
    {\usebeamerfont{author in head/foot}(\insertshortauthor)}%
    \hfill
    {\usebeamerfont{framenumber in head/foot}\textbf{\insertframenumber}}%
  \end{beamercolorbox}%
  \begin{beamercolorbox}[colsep=1.5pt]{lower separation line foot}
  \end{beamercolorbox}
}
% emphasizing means text in default accent color
\renewcommand{\emph}[1]{\textcolor{structure}{#1}}

% \usepackage{microtype} % enable this for improved text typesetting
\usepackage{hyperref} % this is of \href{}{} or \url{}
\usepackage{booktabs} % this is for \midrule \toprule etc in tables


\usetikzlibrary{bayesnet}

\title[Final lecture]{Final lecture}


\begin{document}

\input{../LecturesCommon/titleSlide.tex}




\begin{frame}{Further reading}

\begin{scriptsize}

\begin{itemize}

\item Andrew Gelman, Jennifer Hill, and Masanao Yajima. Why we (usually) don't have to worry about multiple comparisons. Journal of Research on Educational Effectiveness, 5(2), 189-211, 2012.

\item Andrew Gelman, Jessica Hwang, and Aki Vehtari. Understanding predictive
information criteria for Bayesian models. Statistics and Computing, 24(6):
997–1016, 2014.

\item Aki Vehtari, Andrew Gelman, and Jonah Gabry. Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. Statistics and
Computing, 27(5):1413–1432, 2017.

\item John K Kruschke. Bayesian estimation supersedes the t test. Journal of Experimental
Psychology: General, 142(2):573, 2013.

\end{itemize}
\end{scriptsize}
\end{frame}

\begin{frame}{Topics we've covered}

\begin{itemize}
\item Bayesian view of probability.
\item Bayesian learning: Bayes' theorem, prior, likelihood, posterior.
\item Computation: conjugacy, MC, MCMC (semi-conjugacy, Gibbs).
\item MCMC in practice: autocovariance, approximation error, traceplot, effective sample size.
\item Fitting, overfitting.
\item Some standard models: binomial test for proportions, t-test \& ANOVA (equivalent), lm, glm, mixed-effects modelling...
\item Software: Stan, \texttt{rstanarm}, \texttt{mcmcse}...
\end{itemize}

\end{frame}


\begin{frame}{Where do we go from here}

\begin{small}
  \begin{block}{Applying Bayesian statistics as a research tool in your field}
    \begin{itemize}
    \item Kruschke: Doing Bayesian data analysis.
    \item McElreath: Statistical Rethinking: A Bayesian Course with Examples in R and Stan.
    \item Learn R, rstanarm, Stan.
    \item (Gelman...: Bayesian data analysis.)
    \end{itemize}
  \end{block}
  
  \begin{alertblock}{Specializing in Bayesian statistics}
    \begin{itemize}
      \item Ross: First course in probability theory.
      \item Kadane: Principles of uncertainty (subjective Bayesian view).
      \item Robert: The Bayesian choice (objective Bayesian view).
    \end{itemize}

  \end{alertblock}

\end{small}

\end{frame}


\begin{frame}{Topics we've covered}

\begin{itemize}
\item Bayesian view of probability.
\item Bayesian learning: Bayes' theorem, prior, likelihood, posterior.
\item Computation: conjugacy, MC, MCMC (semi-conjugacy, Gibbs).
\item MCMC in practice: autocovariance, approximation error, traceplot, effective sample size.
\item Fitting, overfitting.
\item Some standard models: binomial test for proportions, t-test \& ANOVA (equivalent), lm, glm, mixed-effects modelling...
\item Software: Stan, \texttt{rstanarm}, \texttt{mcmcse}...
\end{itemize}

\end{frame}


\end{document}
